Causal inference methods examples part 2

### 2021-10-29 causal_infernece_examples_v2.Rmd
update

---
title: "Causal inference methods examples part 2"
author: "Ryan Gan"
date: "8/7/2021"
output: html_document
---

```{r setup}
# loading r packages i'll use in this example
library(tmle) # targeted maximum likelihood
library(MASS)
library(xgboost) # gradient boosting with l2 regularization
library(glmnet) # lasso, ridge, and elastic regularization
library(gbm) # gradient boosting machiens
library(shapr) # explainable tree modles
library(tidyverse) # general data wrangling and plotting
library(mgcv) # for non-linear terms
```

@@

- Simple sim
- Interaction and non-linear sim
```{r sim}
# binary confounder example 
set.seed(100)
c1_x_association <- log(5)
c1_y_association <- log(5)
c2_x_association <- log(0.25)
c2_y_association <- log(0.25)

@@

# G-Computation

G-computation (counterfactual estimate) first proposed by Robins in the epi 
literature sometime in the 1990s. In my opinion, this is Pearl's Do-calculus. 
Both state that if we specify a model `outcome | treatment, covariates` sufficiently
to account for confounding bias, then we have achieved condiitonal exchangability.
We can then estimate the marginal effect of treatment on the outcome by taking
the model and predicting as if everyone were treated, and predicting as if 
no one was treated. The average predicted score where everyone is treated can
then be compared to the average predicted score where no one was treated and 
that should be the marginal risk difference (in a binary outcome) or relative
ratio difference.

@@

All covs in model.

```{r g_comp_glm_incorrect}
inc_mod <- glm(outcome ~ trt + c1 + c2 + age + rv_1 + rv_2 + rv_3 + rv_4 + rv_5
               + rv_6 + rv_7 + rv_8 + rv_9 + rv_10, 
               family = binomial(link='logit'), data = df)

# estimate using fit model and take the average of predictions
all_trt_pred_inc <- predict(inc_mod, newdata = all_trt_df, type = 'response') 
no_trt_pred_inc <- predict(inc_mod, newdata = no_trt_df, type = 'response')

# means for risk
trt_mean_inc <- mean( all_trt_pred_inc )
no_trt_mean_inc <- mean( no_trt_pred_inc )

# log odds ratio
mean_log_odds_inc <- mean( 
  log( ( all_trt_pred_inc / (1 - all_trt_pred_inc ) ) / (no_trt_pred_inc / (1 - no_trt_pred_inc)) )
  )

# absolute reduction in risk
print(paste0( 'Risk difference: ', round( trt_mean_inc - no_trt_mean_inc , 2 ) ) )
# relative reduction in risk
print(paste0( 'Relative risk: ', round( trt_mean_inc / no_trt_mean_inc , 2 ) ) )
# odds ratio
gcomp_or_inc <- ( trt_mean_inc / ( 1 - trt_mean_inc ) ) / (no_trt_mean_inc / ( 1 - no_trt_mean_inc ) )
print(paste0('Odds ratio: ', round( gcomp_or_inc , 2)) )

# bias
print(
  paste0('Bias of True - Expected: ', 
         round( trt_out_association_true - mean_log_odds_inc, 2)
         )
  )

@@

Plot showing bias from true
```{r gcomp_bias_plot}
est <- c('Unadjusted GLM', 'Correct Adj GLM', 'G-Comp Adj', 'G-Comp All Cov', 
         'G-Comp Lasso', 'G-Comp XGB')

g_coef <- c(crude_trt_est, adj_mod$coefficients['trt'], log( g_comp_or ), 
            log( gcomp_or_inc), log( gcomp_or_lasso ), log( gcomp_or_xgb ) )

@@

Plots of propensity scores.

@@

```{r lin_att}
ggplot(d) +
  geom_histogram(bins = 50, aes(treatment_p1), alpha = 0.5) + 
  geom_histogram(bins = 50, aes(treatment_p1, weight = w_att, fill = "Treated"), alpha = 0.5) + 
  geom_histogram(bins = 50, alpha = 0.5, aes(x = treatment_p0, y = -..count..)) + 
  geom_histogram(bins = 50, aes(x = treatment_p0, weight = w_att, y = -..count.., fill = "Not Treated"),
                 alpha = 0.5) + 
  scale_fill_manual('', values = c('blue', 'red') ) +
  ylab("Count") + 
  xlab("Propensity") +
  ggtitle('ATT weighting of propensity: linear model') +
  geom_hline(yintercept = 0, lwd = 0.5) +
  scale_y_continuous(label = abs)  +
  theme_linedraw()
```
```{r lin_atm}
ggplot(d) +
  geom_histogram(bins = 50, aes(treatment_p1), alpha = 0.5) + 
  geom_histogram(bins = 50, aes(treatment_p1, weight = w_atm, fill = "Treated"), alpha = 0.5) + 
  geom_histogram(bins = 50, alpha = 0.5, aes(x = treatment_p0, y = -..count..)) + 
  geom_histogram(bins = 50, aes(x = treatment_p0, weight = w_atm, y = -..count.., fill = "Not Treated"),
                 alpha = 0.5) + 
  scale_fill_manual('', values = c('blue', 'red') ) +
  ylab("Count") + 
  xlab("Propensity") +
  ggtitle('ATM weighting of propensity: linear model') +
  geom_hline(yintercept = 0, lwd = 0.5) +
  scale_y_continuous(label = abs)  +
  theme_linedraw()
```

@@

p_score_df <- data.frame(
  estimation = 'Propensity', 
  method = p_est, 
  coef = p_coef
  )  %>% 
  mutate(
    method = factor( method , levels = method ), 
    odds_ratio = exp( coef )
  )

@@

## TMLE

@@

```{r}
summary(tmle_mod)

@@

Final plot summarizing df.

```{r}
ggplot(
  data = final_df, 
  aes( x = factor(method, levels = method), y = odds_ratio , color = estimation) ) +
  geom_point() +
  coord_flip() +
  scale_x_discrete(
    limits = rev(factor(final_df$method, levels = final_df$method))
    )  +
  geom_hline(yintercept = exp( trt_out_association_true ), color = 'red') +
  ylim( 0.25, 1.25 ) +
  theme_linedraw() +
  ylab('Odds Ratio') +
  xlab('Estimation Method')
```

```{r}
  ggplot(data = median_impute_pois %>% filter(!str_detect(term, 'time_factor')), 
       aes( x = factor(term, levels = term), y = log(risk_ratio) ) ) +
  geom_point(position = position_dodge(width = 0.5), 
             color = 'darkblue') +
  geom_errorbar( aes( ymin = log(lower_95), ymax = log(upper_95) ), 
                 width = 0.3, 
                 position = position_dodge(width = 0.5),
                 color = 'darkblue') +
  geom_hline(yintercept = 0, color = 'red', linetype = 'dashed') +
  coord_flip() +
  scale_x_discrete(
    limits = rev(factor(results_table$term, levels = results_table$term))
    ) +
  xlab('Variable (continuous are standardized)') +
  ylab('Log(Risk Ratio)') +
  ggtitle('Poisson model') +
  theme_minimal()
```

>>>>

---
title: "Causal inference methods examples part 2"
author: "Ryan Gan"
date: "2021-10-19"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 2
---

```{r setup}
# loading r packages 
library(tmle) # targeted maximum likelihood
library(MASS)
library(xgboost) # gradient boosting with l2 regularization
library(glmnet) # lasso, ridge, and elastic regularization
library(gbm) # gradient boosting machiens
library(shapr) # explainable tree modles
library(tidyverse) # general data wrangling and plotting
library(mgcv) # for non-linear terms
```

@@

```{r sim}
# binary confounder example 
set.seed(100)
c1_x_association <- log(5)
c1_y_association <- log(5)
c2_x_association <- log(0.25)
c2_y_association <- log(0.25)

@@

# G-computation

G-computation (a.k.a simple substitution, counterfactual estimate) first proposed by Robins in the epi literature sometime in the 1990s. In my opinion, this is Pearl's Do-calculus. Both state that if we specify a model `outcome | treatment, covariates` sufficiently to account for confounding bias, then we have achieved conditional exchangeability. We can then estimate the marginal effect of treatment on the outcome by taking the model and predicting as if everyone were treated, and predicting as if no one was treated. The average predicted score where everyone is treated can then be compared to the average predicted score where no one was treated and that should be the marginal risk difference (in a binary outcome) or relative 6ratio difference.

@@

All covs in model; saturated model.

```{r g_comp_glm_satorrect}
saturated_mod <- glm(outcome ~ trt + c1 + c2 + age + rv_1 + rv_2 + rv_3 + rv_4
                     + rv_5 + rv_6 + rv_7 + rv_8 + rv_9 + rv_10, 
                     family = binomial(link='logit'), data = df)

# estimate using fit model and take the average of predictions
all_trt_pred_sat <- predict(saturated_mod, newdata = all_trt_df, type = 'response') 
no_trt_pred_sat <- predict(saturated_mod, newdata = no_trt_df, type = 'response')

# means for risk
trt_mean_sat <- mean( all_trt_pred_sat )
no_trt_mean_sat <- mean( no_trt_pred_sat )

# log odds ratio
mean_log_odds_sat <- mean( 
  log( ( all_trt_pred_sat / (1 - all_trt_pred_sat ) ) / (no_trt_pred_sat / (1 - no_trt_pred_sat)) )
  )

# absolute reduction in risk
print(paste0( 'Risk difference: ', round( trt_mean_sat - no_trt_mean_sat , 2 ) ) )
# relative reduction in risk
print(paste0( 'Relative risk: ', round( trt_mean_sat / no_trt_mean_sat , 2 ) ) )
# odds ratio
gcomp_or_sat <- ( trt_mean_sat / ( 1 - trt_mean_sat ) ) / (no_trt_mean_sat / ( 1 - no_trt_mean_sat ) )
print(paste0('Odds ratio: ', round( gcomp_or_sat , 2)) )

# bias
print(
  paste0('Bias of True - Expected: ', 
         round( trt_out_association_true - mean_log_odds_sat, 2)
         )
  )

@@

Plot showing bias from true.

```{r gcomp_bias_plot}
est <- c('Unadjusted GLM', 'Correct Adj GLM', 'G-Comp Adj', 'G-Comp All Cov', 
         'G-Comp Lasso', 'G-Comp XGB')

g_coef <- c(crude_trt_est, adj_mod$coefficients['trt'], log( gcomp_or ), 
            log( gcomp_or_sat), log( gcomp_or_lasso ), log( gcomp_or_xgb ) )

@@

Plots of propensity scores weighting schemes.

@@

```{r lin_att}
ggplot(d) +
  geom_histogram(bins = 50, aes(treatment_p1), alpha = 0.5) + 
  geom_histogram(bins = 50, aes(treatment_p1, weight = w_att, fill = "Treated"), alpha = 0.5) + 
  geom_histogram(bins = 50, alpha = 0.5, aes(x = treatment_p0, y = -..count..)) + 
  geom_histogram(bins = 50, aes(x = treatment_p0, weight = w_att, y = -..count.., fill = "Not Treated"),
                 alpha = 0.5) + 
  scale_fill_manual('', values = c('blue', 'red') ) +
  ylab("Count") + 
  xlab("Propensity") +
  ggtitle('ATT weighting of propensity: linear model') +
  geom_hline(yintercept = 0, lwd = 0.5) +
  scale_y_continuous(label = abs)  +
  theme_linedraw()
```

```{r lin_atm}
ggplot(d) +
  geom_histogram(bins = 50, aes(treatment_p1), alpha = 0.5) + 
  geom_histogram(bins = 50, aes(treatment_p1, weight = w_atm, fill = "Treated"), alpha = 0.5) + 
  geom_histogram(bins = 50, alpha = 0.5, aes(x = treatment_p0, y = -..count..)) + 
  geom_histogram(bins = 50, aes(x = treatment_p0, weight = w_atm, y = -..count.., fill = "Not Treated"),
                 alpha = 0.5) + 
  scale_fill_manual('', values = c('blue', 'red') ) +
  ylab("Count") + 
  xlab("Propensity") +
  ggtitle('ATM weighting of propensity: linear model') +
  geom_hline(yintercept = 0, lwd = 0.5) +
  scale_y_continuous(label = abs)  +
  theme_linedraw()
```

@@

p_score_df <- data.frame(
  estimation = 'Propensity', 
  method = p_est, 
  coef = p_coef
  )  %>% 
  mutate(
    method = factor( method , levels = method ), 
    odds_ratio = exp( coef ), 
    percent_bias = coef - log(0.5)
  )

@@

Plot for propensity score estimates difference from the true estimate.

```{r pscore_plot}
ggplot(p_score_df, aes( x = method, y = odds_ratio )) +
  geom_point() +
  geom_hline(yintercept = exp( trt_out_association_true ), color = 'red') +
  ylim( 0.25, 1.25 ) +
  theme_linedraw() +
  ylab('Odds Ratio') +
  xlab('Estimation Method') +
  theme(axis.text.x = element_text(angle = 60, vjust = 0.5))
```

## Targeted-maximum likelihood estimation (TMLE)

@@

```{r tmle_summary}

summary(tmle_mod)

@@

Final plot summarizing bias of estimates compared to true value.

```{r summary_plot}
ggplot(
  data = final_df, 
  aes( x = factor(method, levels = method), y = odds_ratio , color = estimation) ) +
  geom_point() +
  coord_flip() +
  scale_x_discrete(
    limits = rev(factor(final_df$method, levels = final_df$method))
    )  +
  geom_hline(yintercept = exp( trt_out_association_true ), color = 'red') +
  ylim( 0.25, 1.25 ) +
  theme_linedraw() +
  ylab('Odds Ratio') +
  xlab('Estimation Method')
```
